---
title: "CMP713 Project"
author: "Eren Aktug & Ali Burak Erdogan"
date: "1/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparing the data

We load the *Suicide Rates (1985-2016)* data reported by WHO *(see: https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016)* augmented with some additional columns such as Human Development Index (HDI) and GDP (Gross domestic product) values belonging to that country in that year.


```{r, warning=FALSE}
library(readr)
library(tidyr)
library(dplyr)

setwd('/home/burak/Documents/MS/CMP713/proj')
suicide_rates <- read_csv("Suicide Rates (1985-2016).csv")
suicide_rates <- suicide_rates %>% rename(suicides_per_100k = "suicides/100k pop", HDI_for_year = "HDI for year", 
                         gdp_for_year_dollars = "gdp_for_year ($)", gdp_per_capita_dollars = "gdp_per_capita ($)")

```

```{r}
print(suicide_rates)
```

```{r}
summary(suicide_rates)
```

## Filling missing HDI columns using another dataset

Since our dataset consists of 24k rows but 2/3 of HDI index columns are missing (~19k rows) we should fill the missing columns using an additional dataset taken from UN Development Programme *(see: http://hdr.undp.org/en/indicators/137506#)*

```{r, warning=FALSE}
Human_Development_Index_HDI_ <- read_csv("Human Development Index (HDI).csv",
                                         col_types = cols(X4 = col_skip(), X6 = col_skip(),
                                                          X8 = col_skip(), X10 = col_skip(),
                                                          X12 = col_skip(), X14 = col_skip(),
                                                          X16 = col_skip(), X18 = col_skip(),
                                                          X20 = col_skip(), X22 = col_skip(),
                                                          X24 = col_skip(), X26 = col_skip(),
                                                          X28 = col_skip(), X30 = col_skip(),
                                                          X32 = col_skip(), X34 = col_skip(),
                                                          X36 = col_skip(), X38 = col_skip(),
                                                          X40 = col_skip(), X42 = col_skip(),
                                                          X44 = col_skip(), X46 = col_skip(),
                                                          X48 = col_skip(), X50 = col_skip(),
                                                          X52 = col_skip(), X54 = col_skip(),
                                                          X56 = col_skip(), X58 = col_skip(),
                                                          X60 = col_skip(), X62 = col_skip()), skip = 5)

country_HDI <- Human_Development_Index_HDI_[1:189,]
world_avg_HDI <- Human_Development_Index_HDI_[190:207,]

```

```{r}
print(country_HDI)
```

HDI dataset represents each year as a separate column, we need to convert these columns into separate observations using `gather()`

```{r}
country_HDI_tidy <- gather(country_HDI, Year, HDI, "1990":"2019")
country_HDI_tidy$Year <- as.double(country_HDI_tidy$Year)
country_HDI_tidy$HDI <- as.double(country_HDI_tidy$HDI)
country_HDI_tidy <- country_HDI_tidy %>% mutate(Country = ifelse(Country == "Korea (Republic of)", "Republic of Korea", Country),
                                                Country = ifelse(Country == "Czechia", "Czech Republic", Country))
country_HDI_tidy$`HDI Rank` <- NULL
print(country_HDI_tidy)
```

Now we can fill in missing NA values of HDI columns, using our tidied HDI dataset. We perform a left-join operation by matching country and year columns, and observe that rows with missing HDI decreased from 19k+ rows to ~4k rows.
Also, after filling most of HDI cols, we factorize some categorical columns.

```{r}
# join operation and then filling the missing NA values from right table
suicide_rates <- suicide_rates %>% left_join(country_HDI_tidy, by = c("country" = "Country", "year" = "Year")) %>% mutate(HDI_for_year = ifelse(is.na(HDI_for_year), HDI, HDI_for_year)) %>% select(-HDI)

# factorization of some categorical columns
suicide_rates$age <- as.factor(suicide_rates$age)
suicide_rates$country <- as.factor(suicide_rates$country)
suicide_rates$sex <- as.factor(suicide_rates$sex)
suicide_rates$generation <- as.factor(suicide_rates$generation)
suicide_rates$`country-year` <- NULL

summary(suicide_rates)
```



