---
title: "CMP713 Project"
author: "Eren Aktug & Ali Burak Erdogan"
date: "1/8/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preparing the data

We load the *Suicide Rates (1985-2016)* data reported by WHO *(see: https://www.kaggle.com/russellyates88/suicide-rates-overview-1985-to-2016)* augmented with some additional columns such as Human Development Index (HDI) and GDP (Gross domestic product) values belonging to that country in that year.


```{r, warning=FALSE}
library(readr)
library(tidyr)
library(dplyr)

setwd('/home/burak/Documents/MS/CMP713/proj')
suicide_rates <- read_csv("Suicide Rates (1985-2016).csv")
suicide_rates <- suicide_rates %>% rename(suicides_per_100k = "suicides/100k pop", HDI_for_year = "HDI for year", 
                         gdp_for_year_dollars = "gdp_for_year ($)", gdp_per_capita_dollars = "gdp_per_capita ($)")

```

```{r}
print(suicide_rates)
```

```{r}
summary(suicide_rates)
```

# Data Preprocessing

Since our dataset consists of 24k rows but 2/3 of HDI index columns are missing (~19k rows) we should fill the missing columns using an additional dataset taken from UN Development Programme *(see: http://hdr.undp.org/en/indicators/137506#)*
*(Also see how HDI calculated: http://hdr.undp.org/sites/default/files/hdr2018_technical_notes.pdf )*

```{r, warning=FALSE}
Human_Development_Index_HDI_ <- read_csv("Human Development Index (HDI).csv",
                                           col_types = cols(`1990` = col_double(), `1991` = col_double(), `1992` = col_double(), `1993` = col_double(), `1994` = col_double(), `1995` = col_double(), `1996` = col_double(), `1997` = col_double(), `1998` = col_double(), `1999` = col_double(), `2000` = col_double(), `2001` = col_double(), `2002` = col_double(), `2003` = col_double(), `2004` = col_double(), `2005` = col_double(), `2006` = col_double(), `2007` = col_double(), `2008` = col_double(), `2009` = col_double(), `2010` = col_double(), `2011` = col_double(), `2012` = col_double(), `2013` = col_double(), `2014` = col_double(), `2015` = col_double(), `2016` = col_double(), `2017` = col_double(), `2018` = col_double(), `2019` = col_double(),
                                                          X4 = col_skip(), X6 = col_skip(),
                                                          X8 = col_skip(), X10 = col_skip(),
                                                          X12 = col_skip(), X14 = col_skip(),
                                                          X16 = col_skip(), X18 = col_skip(),
                                                          X20 = col_skip(), X22 = col_skip(),
                                                          X24 = col_skip(), X26 = col_skip(),
                                                          X28 = col_skip(), X30 = col_skip(),
                                                          X32 = col_skip(), X34 = col_skip(),
                                                          X36 = col_skip(), X38 = col_skip(),
                                                          X40 = col_skip(), X42 = col_skip(),
                                                          X44 = col_skip(), X46 = col_skip(),
                                                          X48 = col_skip(), X50 = col_skip(),
                                                          X52 = col_skip(), X54 = col_skip(),
                                                          X56 = col_skip(), X58 = col_skip(),
                                                          X60 = col_skip(), X62 = col_skip()), skip = 5)

country_HDI <- Human_Development_Index_HDI_[1:189, 2:32]
development_avg_HDI <- Human_Development_Index_HDI_[191:194, 2:32]
development_avg_HDI$Country <- as.factor(development_avg_HDI$Country)
names(development_avg_HDI)[names(development_avg_HDI) == "Country"] <- "development" # rename column
levels(development_avg_HDI$development) <- c("high", "low", "medium", "veryHigh")
world_avg_HDI <- Human_Development_Index_HDI_[206, 3:32]
```

HDIs for each country across the years:
```{r}
print(country_HDI)
```

HDI cutoff points for development status across each year:
```{r}
print(development_avg_HDI)
```

# Generating "Country Development Status" column 

We are identifying development status of each country for each year.

Since we observe a constant increase in the Human Development Index all over the world as the years pass, we need a metric to indicate that how much developed a country was in a given year, and how much was the HDI index compared to global average.

That's why we use the additional information given by HDI dataset which shows the average HDI values of all countries and continents.

We label each row with development category (i.e. the development category belongs to a country in a specific year)

```{r}
years <- colnames(country_HDI[, -1])
country_HDI_development_labels <- country_HDI

# process each year column 
for(year in years) {
  tmp_HDIs <- unlist(country_HDI_development_labels[year])
  tmp_breaks <- development_avg_HDI[3:1, year]
  #country_HDI_development_labels[year] <- cut(tmp_HDIs, breaks = c(0, tmp_breaks, 1),  labels=development_avg_HDI[, "development"])
  country_HDI_development_labels[year] <- cut(tmp_HDIs, breaks = c(0, unlist(tmp_breaks), 1), labels = rev(unlist(development_avg_HDI[, "development"])) )
}
rm(tmp_HDIs, tmp_breaks. year, years)

country_HDI_development_labels <- gather(country_HDI_development_labels, Year, Development, "1990":"2019")
country_HDI_development_labels$Year <- as.double(country_HDI_development_labels$Year)

```

Resulting table

```{r}
print(country_HDI_development_labels)
```


HDI dataset represents each year as a separate column, we need to convert these columns into separate observations using `gather()`

```{r}
country_HDI_tidy <- gather(country_HDI, Year, HDI, "1990":"2019")
country_HDI_tidy$Year <- as.double(country_HDI_tidy$Year)
country_HDI_tidy$HDI <- as.double(country_HDI_tidy$HDI)
country_HDI_tidy <- country_HDI_tidy %>% mutate(Country = ifelse(Country == "Korea (Republic of)", "Republic of Korea", Country),
                                                Country = ifelse(Country == "Czechia", "Czech Republic", Country))

print(country_HDI_tidy)
```

And we are joining the previously prepared development labels table for each country in each year:

```{r}
country_HDI_tidy <- country_HDI_tidy %>% left_join(country_HDI_development_labels, by = c("Country", "Year"))
print(country_HDI_tidy)
```



## Filling missing HDI columns using our separate HDI data

Now we can fill in missing NA values of HDI columns, using our tidied HDI dataset. We perform a left-join operation by matching country and year columns, and observe that rows with missing HDI decreased from 19k+ rows to ~4k rows.
Also, after filling most of HDI cols, we factorize some categorical columns.

```{r}
# join operation and then filling the missing NA values from right table
suicide_rates <- suicide_rates %>% left_join(country_HDI_tidy, by = c("country" = "Country", "year" = "Year")) %>% mutate(HDI_for_year = ifelse(is.na(HDI_for_year), HDI, HDI_for_year)) %>% select(-HDI)

# factorization of some categorical columns
suicide_rates$age <- as.factor(suicide_rates$age)
suicide_rates$country <- as.factor(suicide_rates$country)
suicide_rates$sex <- as.factor(suicide_rates$sex)
suicide_rates$generation <- as.factor(suicide_rates$generation)
suicide_rates$`country-year` <- NULL
names(suicide_rates)[names(suicide_rates) == "Development"] <- "country_development"

summary(suicide_rates)
```



